The evolution of gene expression in the term placenta of viviparous mammals

### 2016-09-01 presentation/placenta_viviparous_presentation.Rnw
add core placenta shape plot

\documentclass[ignorenonframetext]{beamer}
\usepackage{morewrites}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{fancybox}
\usepackage{booktabs}
\usepackage{pdfpcnotes}
% \usepackage{multirow}
\usepackage{acronym}
\usepackage{qrcode}
\usepackage{cleveref}
\usepackage[absolute,overlay]{textpos}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature]{biblatex}
\addbibresource{references.bib}

\usepackage{texshade}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usepackage{xspace}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\renewcommand*{\bibfont}{\tiny}
\usepackage{mathtools,cancel}
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
\AfterEndEnvironment{minted}{\end{tcolorbox}}%

@@

<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
    source(a)
}
@

@@

\begin{frame}{Species Studied}
% \begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
\end{figure}
\end{frame}

@@

\subsection{Using Make}

\begin{frame}{Makefile rules}
\centering
\includegraphics[keepaspectratio,height=0.8\textheight]{figures/default}
\end{frame}

\begin{frame}[fragile]{Common Options}
\tiny
\inputminted[firstline=8, lastline=36]{make}{../fastq/common_makefile}
\end{frame}

\begin{frame}[fragile]{FASTQC}
\inputminted[firstline=78, lastline=83]{make}{../fastq/common_makefile}
\end{frame}

\begin{frame}[fragile]{Trinity}
\tiny
\inputminted[firstline=211, lastline=222]{make}{../fastq/common_makefile}
\end{frame}

\begin{frame}[fragile]{Diamond}
\tiny
\inputminted[firstline=249, lastline=268]{make}{../fastq/common_makefile}
\end{frame}

\begin{frame}[fragile]{STAR Indexes}
\tiny
\inputminted[firstline=86, lastline=105]{make}{../fastq/common_makefile}
\end{frame}

\begin{frame}[fragile]{STAR}
\tiny
\inputminted[firstline=106, lastline=118]{make}{../fastq/common_makefile}
\end{frame}

\begin{frame}[fragile]{Cufflinks}
\tiny
\inputminted[firstline=43, lastline=60]{make}{../fastq/common_makefile}
\end{frame}

@@

>>>>

\documentclass[ignorenonframetext]{beamer}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{fancybox}
\usepackage{booktabs}
\usepackage{pdfpcnotes}
% \usepackage{multirow}
\usepackage{acronym}
\usepackage{qrcode}
\usepackage{cleveref}
\usepackage[absolute,overlay]{textpos}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature]{biblatex}
\addbibresource{references.bib}

\usepackage{texshade}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usepackage{xspace}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\renewcommand*{\bibfont}{\tiny}
\usepackage{mathtools,cancel}

@@

<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R","plot_ancestral_state_tree.R")))>>=
for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R",
                                       "utility_functions.R",
                                       "plot_ancestral_state_tree.R"
                                       ))) {
    source(a)
}
@

@@

\begin{frame}{Species Studied}
\begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
\end{figure}
\end{frame}

@@

@@

\begin{frame}{Core Placenta Transcriptome Shape}
\begin{figure}
<<core.placenta.transcriptome.distribution,fig.width=6,fig.height=4,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
num.species <- star.logs.wide[,length(unique(species))]
calculate.distribution.shape <- 
    function(min.vector) {
        min.vector <- sort(min.vector,decreasing=TRUE)
        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
        return(temp)
    }

cptg.nh <- core.placenta.transcriptome.genes.shape[!(human_name %in% housekeeping.genes.superset[,gene_short_name])]

cpt.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression"),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species," species"),
              x
              )
    nc.levels <- paste0(num.species," of ",num.species," species")
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species," species"),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species," species"))
    }
    return(factor(x,levels=nc.levels))
}

cpt.distribution.shape <- 
    rbindlist(cpt.distribution.shape)

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Minimum Expression in`:=
                                fix.expression.names(`Minimum Expression in`)]

setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")

print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
             aes(x=`Expression (FPKM)`,color=`Minimum Expression in`,y=num_genes))
      + geom_line(lwd=1)
      + xlab("Minimum Expression cutoff (FPKM)")
      + ylab("Genes with N-th Minimum Expression")
      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + geom_vline(xintercept=10)
      + geom_segment(mapping=aes(y=V1,
                         yend=V1,
                         xend=10,
                         x=1000,
                         color=`Minimum Expression in`),
                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Minimum Expression in`],
                     arrow=arrow(length=unit(0.5,"cm"))
                   )
      + geom_text(mapping=aes(x=1000,
                      y=V1,
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Minimum Expression in`],
                  show.legend=FALSE,
                  hjust=0
                  )
      + guides(color=guide_legend(title="Minimum\nExpression in"))
      )
@ 
\end{figure}
\end{frame}
